---
title: "`allocation`: Exact Optimal Allocation Algorithms for Stratified Sampling"
author:
  - name: Andrew M. Raim
    email: andrew.raim@census.gov
  - name: Tommy Wright
    email: Tommy.Wright@census.gov
affiliations:
  - name: U.S. Census Bureau
    department: Center for Statistical Research and Methodology
    address: 4600 Silver Hill Road
    city: Washington DC
    country: U.S.A.
format:
  pdf:
    fontsize: 10pt
    indent: false
    toc: true
    number-sections: true
    colorlinks: true
    link-citations: true
    prompt: false
    include-in-header:
      text: |
        \usepackage{common}
vignette: >
  %\VignetteIndexEntry{allocation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::pdf}
bibliography: references.bib
editor_options: 
  chunk_output_type: console
abstract: |
  Abstract goes here ...
thanks: |
  corresponds to `allocation` version `{r} packageVersion("allocation")`.
  Methodology, U.S. Census Bureau, Washington, DC, 20233, U.S.A.
  Document was compiled `{r} format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")` and
  **Contact**`:` <andrew.raim@census.gov>, Center for Statistical Research &
geometry:
  - left=0.75in
  - right=0.75in
  - top=0.75in
  - bottom=1.00in
code-block-bg: "#FAFAFA"
code-block-border-left: "#008CFF"
callout-icon: false
filters:
  - include-code-files
execute:
  eval: true
---

```{r}
#| include: false
library(allocation)
set.seed(1234)
```

# Disclaimer and Acknowledgments {-}

This document is released to inform interested parties of ongoing research and
to encourage discussion of work in progress. Any views expressed are those of
the authors and not those of the U.S. Census Bureau.

Although there are no guarantees of correctness of the `allocation` package,
reasonable efforts will be made to address shortcomings. Comments, questions,
corrections, and possible improvements can be communicated through the
Github [repository](https://github.com/andrewraim/allocation) for the package.

# Introduction {#sec-intro}

TBD: @Neyman1934, @Wright2012, and @Wright2017

TBD: variance target

```{r}
#| message: false
#| warning: false
library(Rmpfr)
```

# Overview of the Package

TBD

# Examples

## Allocation for Fixed Sample Size

Run Algorithm III using an example in @Wright2017.

```{r}
N = c(47, 61, 41)
S = sqrt(c(100, 36, 16))
lo = c(1,2,3)
hi = c(5,6,4)
n0 = 10

out1 = allocate_fixn(n0, N, S, lo, hi)
print(out1)
```

To see details justifying each selection, run `allocate_fixn` with the option
`verbose = TRUE`.

Compare the above results to Neyman allocation.

```{r}
out2 = allocate_neyman(n0, N, S)
print(out2)
```

Internally, we work with high precision numbers via the `Rmpfr` package.
We provide an `allocation` accessor function to extract the allocation as a
numeric vector.

```{r}
allocation(out1)
allocation(out2)
```

The numerical precision and number of decimal points printed, can be
changed by setting a global option for the `allocation` package.

```{r}
options(allocation.prec.bits = 256)
options(allocation.print.decimals = 4)
```

## Allocation for Target Precision

Run Algorithm IV using an example in @Wright2017. Since our target
variance `v0` is a very large number, we pass it as an `mpfr` object to
avoid loss of precision.

```{r}
H = 10
v0 = mpfr(388910760, 256)^2
N = c(819, 672, 358, 196, 135, 83, 53, 40, 35, 13)
lo = c(3,3,3,3,3,3,3,3,3,13)
S = c(330000, 518000, 488000, 634000, 1126000, 2244000, 2468000, 5869000,
	29334000, 1233311000)

out1 = allocate_prec(v0, N, S, lo)
print(out1)
```

To see details justifying each selection, run `allocate_prec` with the
option `verbose = TRUE`.

Compare the above results to Neyman allocation. Here, we first need to
compute a target sample size. This is done with a given cv and revenue
data. See @Wright2017 for details. We also exclude the 10th stratum
from the allocation procedure, as it is a certainty stratum; its
allocation is considered fixed at 13.

```{r}
cv = 0.042
rev = mpfr(9259780000, 256)
n = sum(N[-10] * S[-10])^2 / ((cv * rev)^2 + sum(N[-10] * S[-10]^2))
out2 = allocate_neyman(n, N[-10], S[-10])
print(out2)
```

Extract the final allocations.

```{r}
allocation(out1)
allocation(out2)
```

# References {-}

::: {#refs}
:::
