---
title: "Demonstration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(Rmpfr)
library(allocation)
```

# Algorithm III: Sampling with Target Sample Size
Test Algorithm III using the example in the Wright (2017).

```{r}
N.str <- c(47, 61, 41)
S.str <- sqrt(c(100, 36, 16))
lo.str <- c(1,2,3)
hi.str <- c(5,6,4)
n <- 10

out1 <- algIII(n, N.str, S.str, lo.str, hi.str, verbose = TRUE)
print(out1)
```

Compare the above results to Neyman allocation
```{r}
out2 <- neyman(n, N.str, S.str, verbose = TRUE)
print(out2)
```

Internally, we work with high precision numbers via the `Rmpfr` package. We provided an `alloc` accessor function to extract the allocation as a numeric vector.

```{r}
alloc(out1)
alloc(out2)
```

The numerical precision and number of decimal points printed, can be changed by setting a global option for the `allocation` package.
```{r}
options(allocation.prec.bits = 256)
options(allocation.print.decimals = 4)
```

# Algorithm IV: Sampling with Target Variance
Test Algorithm IV using the example in the Wright (2017). Since our target variance `v0` is a very large number, we pass it as an `mpfr` object to avoid loss of precision.
```{r}
H <- 10
v0 <- mpfr(388910760, 256)^2
N.str <- c(819, 672, 358, 196, 135, 83, 53, 40, 35, 13)
lo.str <- c(3,3,3,3,3,3,3,3,3,13)
S.str <- c(330000, 518000, 488000, 634000, 1126000, 2244000, 2468000, 5869000, 29334000, 1233311000)

out1 <- algIV(v0, N.str, S.str, lo.str, verbose = TRUE)
print(out1)
```

Compare the above results to Neyman allocation. Here, we first need to compute a target sample size. This is done with a given cv and revenue data. See Wright (2017) for details. We also exclude the 10th stratum from the allocation procedure, as it is a certainty stratum; its allocation is considered fixed at 13.
```{r}
cv <- 0.042
rev <- mpfr(9259780000, 256)
n <- sum(N.str[-10] * S.str[-10])^2 / ((cv * rev)^2 + sum(N.str[-10] * S.str[-10]^2))
out2 <- neyman(n, N.str[-10], S.str[-10], verbose = TRUE)
print(out2)
```

Extract the final allocations.
```{r}
alloc(out1)
alloc(out2)
```

```