---
title: "Demonstration"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Rmpfr)
library(allocation)
```

# Overview
This package implements several exact optimization methods to allocate samples to strata. See the following references for details.

1. Tommy Wright (2012). The Equivalence of Neyman Optimum Allocation for
Sampling and Equal Proportions for Apportioning the U.S. House of
Representatives. The American Statistician, 66, pp.217-224.
2. Tommy Wright (2017), Exact optimal sample allocation: More efficient than
Neyman, Statistics & Probability Letters, 129, pp.50-57.

# Installation
To install from Github, obtain the `devtools` package and run the following in R. The `Rmpfr` package is used - both inside the `allocation` package and to set up the following examples - for high precision arithmetic.

```{r, eval=FALSE}
install.packages("Rmpfr")
devtools::install_github("andrewraim/allocation")
```

# Usage

```{r, eval=FALSE}
library(Rmpfr)
library(allocation)
```

## Algorithm III: Sampling with Target Sample Size
Run Algorithm III using an example in Wright (2017).

```{r}
N_str = c(47, 61, 41)
S_str = sqrt(c(100, 36, 16))
lo_str = c(1,2,3)
hi_str = c(5,6,4)
n = 10

out1 = algIII(n, N_str, S_str, lo_str, hi_str)
print(out1)
```
To see details justifying each selection, run `algIII` with the option `verbose = TRUE`.

Compare the above results to Neyman allocation
```{r}
out2 = neyman(n, N_str, S_str)
print(out2)
```

Internally, we work with high precision numbers via the `Rmpfr` package. We provided an `alloc` accessor function to extract the allocation as a numeric vector.

```{r}
alloc(out1)
alloc(out2)
```

The numerical precision and number of decimal points printed, can be changed by setting a global option for the `allocation` package.
```{r}
options(allocation.prec.bits = 256)
options(allocation.print.decimals = 4)
```

## Algorithm IV: Sampling with Target Variance
Run Algorithm IV using an example in Wright (2017). Since our target variance `v0` is a very large number, we pass it as an `mpfr` object to avoid loss of precision.
```{r}
H = 10
v0 = mpfr(388910760, 256)^2
N_str = c(819, 672, 358, 196, 135, 83, 53, 40, 35, 13)
lo_str = c(3,3,3,3,3,3,3,3,3,13)
S_str = c(330000, 518000, 488000, 634000, 1126000, 2244000, 2468000, 5869000, 29334000, 1233311000)

out1 = algIV(v0, N_str, S_str, lo_str)
print(out1)
```
To see details justifying each selection, run `algIV` with the option `verbose = TRUE`.

Compare the above results to Neyman allocation. Here, we first need to compute a target sample size. This is done with a given cv and revenue data. See Wright (2017) for details. We also exclude the 10th stratum from the allocation procedure, as it is a certainty stratum; its allocation is considered fixed at 13.
```{r}
cv = 0.042
rev = mpfr(9259780000, 256)
n = sum(N_str[-10] * S_str[-10])^2 / ((cv * rev)^2 + sum(N_str[-10] * S_str[-10]^2))
out2 = neyman(n, N_str[-10], S_str[-10])
print(out2)
```

Extract the final allocations.
```{r}
alloc(out1)
alloc(out2)
```

```